# Reproducible Research: Peer Assessment 1
----  

```{r setup,echo=FALSE}
library(markdown)
library(knitr)
library(plyr)
library(stringr)
library(ggplot2)

setwd("~/RepData_PeerAssessment1")
temp <- unzip("activity.zip")
```
----   

## Loading and preprocessing the data  
**Rubric:  Is there code to read the data?**   
```{r readdata,echo=TRUE}

df <- read.csv(file = "activity.csv"
               , head=TRUE
               , na.strings = "NA"
               )
```
**Rubric:  There is code presented for reading in the dataset**  
I want a datetime column...   

```{r makedatetime, echo=TRUE}
df$datetime<-strptime(paste(df$date
                                , " "
                                 , str_sub(paste("0",(60*df$interval%/%100)/60,sep=""),-2,-1)
                                 , ":"           
                                 , str_sub(paste("0",df$interval%%100,sep=""),-2,-1)
                                , ":00"
                                 , sep=""
                           )
                     , "%Y-%m-%d %H:%M:%S"
                     )
df$datetime<- as.POSIXct(strptime(as.character(df$datetime)
                                  , "%Y-%m-%d %H:%M:%S"
                                  )
                         )
```
... and a Weekday/Weekend factor...  

```{r makeweekend,echo=TRUE}
df$weekend <- as.factor(
                        weekdays(df$datetime)=="Saturday"| 
                        weekdays(df$datetime)=="Sunday"
                        )
df$weekend <- ordered(df$weekend,
                     levels = c(TRUE,FALSE),
                     labels = c("Weekend", "Weekday"))
```

Check...  

```{r takealook,echo=TRUE}
str(df)
head(df)
```

----  

## What is mean total number of steps taken per day?  
**Rubric: There is a histogram of the total number of steps taken each day**  

```{r makehist1,echo=TRUE, fig.height=4, fig.width=8}

StepsPerDay <-aggregate(steps~date, data=df, sum, na.rm=TRUE)
hist(StepsPerDay$steps
     , xlab="Steps Per Day"
     , main= "Missing Values Removed"
     , cex.axis=.8
     , labels=TRUE
     , ylim=c(0, 40)
     , col="light blue"
     )
grid(NA,NULL,lwd = 2) 
box()
```
Add up the steps for each date...  

```{r makestepsperday,echo=TRUE}

StepsPerDay <-aggregate(steps~date, data=df, sum, na.rm=TRUE)
head(StepsPerDay)
```

**Rubric: Both the mean and median are reported.**  
```{r,stepstats,echo=TRUE}
mean(StepsPerDay$steps)
median(StepsPerDay$steps)
```
  
----  

##What is the average daily activity pattern?  
**Rubric: A time series plot is included and it appears correct.**  
```{r maketsplot,echo=TRUE, fig.height=4, fig.width=8}
AvgStepsPerInterval <- aggregate(steps~interval, data=df,mean,na.rm=TRUE)

MaxPt <-AvgStepsPerInterval[AvgStepsPerInterval$steps==
                                    max(AvgStepsPerInterval$steps),]

plot(  AvgStepsPerInterval$interval
     , AvgStepsPerInterval$steps
     , type='l' 
     , xlab="Interval"
     , ylab= "Average # Steps"
     , col="blue"
    )
text(  MaxPt$interval
     , MaxPt$steps
     , paste("Max =", trunc(MaxPt$steps,0), "steps @ interval", MaxPt$interval)
     , cex = .8
     , pos = 4
     , col = "red"
    )
grid(NULL,NULL,lwd = 2) 
```
  
**Rubric: The 5-minute interval containing the maximum number of steps is reported**  

Here's the answer: The 5-minute interval containing the maximum number of steps is `r MaxPt$interval` and the maximum # of steps is `r MaxPt$steps`, code is shown below...  

```{r maximum1,echo=TRUE}
MaxPt
```

----  

## Imputing missing values
Find out how many missing values are in each column of the data...  

```{r summarize,echo=TRUE}
nrow(df)-nrow(df[complete.cases(df),])
```

### 2. Devise a strategy for filling in all of the missing values in the dataset. 
**Rubric: There is a description of a strategy for imputing missing data.**  

Description:  My strategy is to replace each missing value with the average # of steps during the same 5-minute interval from the days non-missing values for that interval.  

Please see the code chunk below for the calculation.  

NOTE:  I found this code [here](http://stackoverflow.com/questions/9322773/how-to-replace-na-with-mean-by-subset-in-r-impute-with-plyr).   

The new dataset will be named df.imp as shown in the code chunk below.  

**Rubric: The code for imputing missing data is shown.**  

```{r imputena,echo=TRUE}
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
df.imp <- ddply(df, ~ interval, transform, steps = impute.mean(steps))  
df.imp<-df.imp[order(df.imp$date,df.imp$interval), ] 
```

Now compare the two data frames to verify no more NAs (this should give zero)...  

```{r newsummarize,echo=TRUE}
nrow(df.imp)-nrow(df.imp[complete.cases(df.imp),])
```
Yes, missing values are gone.  

**Rubric: the histogram is present in the report, for after missing values were imputed.**  

```{r hist2, echo=TRUE, fig.height=4, fig.width=8}
  
StepsPerDay.Imp <-aggregate(steps~date, data=df.imp, sum)
hist(  StepsPerDay.Imp$steps
     , xlab = "Steps Per Day"
     , main = "Missing Values Imputed"
     , cex.axis = .8
     , labels = TRUE
     , ylim = c(0, 40)
     , col = "light blue"
     )
grid(NA,NULL,lwd = 2) 
box()
```

Observatoion:  The histogram is slightly changed compared to the first one, with more days falling into the 10-15K bin.  

## What is the impact of imputing missing data on the estimates of the total daily number of steps?  

Just for further comparison, I also compared the statistics.  

```{r stats2,echo=TRUE}

c(sum(StepsPerDay$steps,na.rm=TRUE),sum(StepsPerDay.Imp$steps,na.rm=TRUE))
```
The sum of the steps increased due to imputing missing values, as expected.  

```{r stats3,echo=TRUE}
c(mean(StepsPerDay$steps,na.rm=TRUE),mean(StepsPerDay.Imp$steps,na.rm=TRUE))
```
No change to the average per day, thanks to imputing using averages.  

```{r stats4,echo=TRUE}
c(median(StepsPerDay$steps,na.rm=TRUE),median(StepsPerDay.Imp$steps,na.rm=TRUE))
```
No change to the median steps per day.  

```{r newmax,echo=TRUE}
MaxPt.Imp
```
No change to the interval in which the maximum occurred.  

The mean, median and maximum stayed almost exactly the same.  The apparent impact of imputing missing values using interval averages was to add more steps in the middle (10-15K per day) range, as seen on the histogram.  

 ----  
 
## Are there differences in activity patterns between weekdays and weekends?  

**NOTE TO ASSESSOR: The code to create the weekend factor is at top of this page in the preprocessing section.**  

**Rubric: There is a panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends. (with missing values imputed.)**  

```{r panelplot,echo=TRUE, fig.height=4, fig.width=8}
AvgStepsPerIntervalWkend.Imp <- aggregate(steps~interval+weekend, data=df.imp,mean)
sp <- ggplot(AvgStepsPerIntervalWkend.Imp, aes(x=interval, y=steps)) + geom_line()+ ylab("Number of Steps")
sp + facet_wrap(~weekend, ncol=1)
```
Conclusion:  It looks like the person's weekday activity was greater in the mornings, while weekend activity was more spread out throughout the day, on average.   

** Rubric:  The report contains R code for all of the results presented.**  

Dear Peer Assessors, I tried to keep it simple for you by sticking to the rubric.  Please leave a comment if you deducted any points.  Thanks!  

# The End



